Material Methods

### 2020-04-17 doc/MatMeth.Rmd
starting new outline for material and methods

@@

The mean, variance and coefficient of variance for MLH1 counts per cell were calculated for each mouse (in R).

**M1. Mixed Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex + rand(strain) + \varepsilon $$

**M2. Linear Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex * strain + \varepsilon$$

**M3. Linear Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ sex * strain + \varepsilon$$

**M4. Linear Model** $$sex\ specific\ mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * strain + \varepsilon$$

We chose to use a **mixed model framework** to interpret the patterns of variation, built using the lmer4 package (cite) in R (v3.5.2)[@Rstudio]. (lme)

@@

@@

@@

The total SC per cell and single bivalent measures were generated using the approaches discribed in [@wang_2019_sc] and [@peterson2019] respectively.

Hand measures of SC/pactyene / meiotic chromosomes generated by using ImageJ/Fiji (v1.52) [@fiji].

to quantify / charaecterize the SC (length) (synapsed chromosomes, )-- two image analysis algorithms were used both which apply a 'skeletonizing' transformation to the synapsed chromosomes,  which transforms the elongatde chromosome space into a single pixel wide 'trace' of the originial chromosom shape.

The DNA CrossOver script isolated single chromosomes shapes from an image, in addition to returning the SC length -- also measured the location of (green and blue signal along the single SC - reflecting MLH1 foci and cetromere signal respectively).

**For SC based metric**

- To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a reduced dataset including only bivalents with SC lengths below the 2nd quartile (for SC length) in each cell. This dataset included the four or five shortest bivalents  and excluded the X bivalent in oocytes. A total of 678 -- `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from 103 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and 37 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes. Although this smaller dataset has decreased power, it offers a more comparable set of single bivalents to compare between the sexes.

- As an additional metric of chromatin compaction, we computed the total (summed) SC length of all bivalents for single cells, using the image analysis algorithm of [@wang_2019_sc]. Mouse means were calculated from cell-wide total SC lengths in 2,984 -- `r length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` out of 3,680 -- `r length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` cells with MLH1 counts (Figure X).

@@

The single bivalent data set was curated of algorithm mistakes manually>. For charaecterizing any of the foci positions or interfocal distances any normalized refers to dividing the raw measures by the total SC for that invidiualy bivalent.

**chromosome size effects**

@@

< manual measures of whole cell bivalent measures (measures of all bivalents were incorporated into this data set. ), were compiled by mixed automated bivalent measures with hand measurements. (Since the error in automated measures was so low)>

@@

- This metric is closer to measures of interference from linkage maps which are also removed from physical scales in that they measure frequencies of crossovers. Crossover interference is stronger in male specific linkage maps compared to female maps in a variety of species (ref).  **IFD^norm^**

**CO Interferecne**

Our primary mode of examining interference is to compared the interfocal disatnces (IFDs) (raw and normalized by SC length.)
Gamma parameter v has historically been used to charaecterize interference and was calculated using <x function>.

>>>>

#MLH1 + Bivalent

#-models (how several were used)
#the code of models
#the tests / analysis of models

@@

<The mean, variance and coefficient of variance for MLH1 counts per cell were calculated for each mouse (in R).>

@@

We applied the following models to analze several dependant variables: mouse average mlh1 count per cell, the mouse average of variance in crossover count per cell, mouse mean SC length, mouse mean IFD^norm^ and IFD^raw^, and X.

**M1. Mixed Model** $$mouse \ av.\ dependant\ variable\ ~=~ subsp * sex + rand(strain) + \varepsilon $$

**M2. Linear Model** $$mouse \ av.\ dependant\ variable\ ~=~ subsp * sex * strain + \varepsilon$$

**M3. Linear Model** $$mouse \ av.\ dependant\ variable\ ~=~ sex * strain + \varepsilon$$

**M4. Linear Model** $$sex\ specific\ mouse \ av.\ dependant\ variable\ ~=~ subsp * strain + \varepsilon$$

@@

We chose to use a **mixed model framework** to interpret the patterns of variation, built using the lmer4 package (cite) in R (v3.5.2)[@Rstudio]. <We used the lme package - for estimating random variabkes>
<We used x package for testing significance of fixed effects and V package for testing significance of (strain) random variables (strain)>

@@

# image analysis programs
# Richard total sc

# DNA.CrossOver (site, efficiency, curation, chromosome size effects-reduced dataset,

# chromosome proportions

# sc bivalent used in models (aldready above)

# IFD / interference
```

**CO Interferecne**  -- Our primary mode of examining interference is to compared the interfocal disatnces (IFDs) (raw and normalized by SC length.)  -- Gamma parameter v has historically been used to charaecterize interference and was calculated using <x function>.

<to quantify / charaecterize the SC (length) (synapsed chromosomes, )-- two image analysis algorithms were used both which apply a 'skeletonizing' transformation to the synapsed chromosomes,  which transforms the elongatde chromosome space into a single pixel wide 'trace' of the originial chromosom shape.>

To quantify the SC area of indiviual cells and single chromosomes we applied the approaches discribed in [@wang_2019_sc] and [@peterson2019] respectively.  Both which apply a 'skeletonizing' transformation to the synapsed chromosomes,  which transforms the elongatde chromosome space into a single pixel wide 'trace' of the originial chromosom shape. Hand measures of SC/pactyene / meiotic chromosomes generated by using ImageJ/Fiji (v1.52) [@fiji].

##### total SC - Wang

As an additional metric of chromatin compaction, we computed the total (summed) SC length of all bivalents for single cells, using the image analysis algorithm of [@wang_2019_sc]. Mouse means were calculated from cell-wide total SC lengths in 2,984 -- `r length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` out of 3,680 -- `r length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` cells with MLH1 counts (Figure X). the mouse average total sc was applied to the model framework.  <how else was this analyzed?!>

##### DNACrossover

The DNACrossover algorithm is hosted on CyVerse, more detailed description in @peterson2019. The DNA CrossOver algorithm isolated single chromosomes shapes from an image, in addition to returning the SC length and also measured the location of (green and blue signal along the single SC - reflecting MLH1 foci and cetromere signal respectively).

**CURATION data ect** (i think these are still in results)

@@

**REDUCED DATASET**

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a reduced dataset including only bivalents with SC lengths below the 2nd quartile (for SC length) in each cell. This dataset included the four or five shortest bivalents  and excluded the X bivalent in oocytes. A total of 678 -- `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from 103 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and 37 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes. Although this smaller dataset has decreased power, it offers a more comparable set of single bivalents to compare between the sexes.

@@

**chromosome proportion**

@@

**IFD, Interference**

In order to examine interference, Interfocal distance was measured from the single bivalent data sets.
- This metric is closer to measures of interference from linkage maps which are also removed from physical scales in that they measure frequencies of crossovers. Crossover interference is stronger in male specific linkage maps compared to female maps in a variety of species (ref).  **IFD^norm^**
